\setchapterpreamble[or][.5\textwidth]{%
\dictum[GLaDOS]{%
This was a triumph.\\
I'm making a note here: HUGE SUCCESS.\\
It's hard to overstate my satisfaction.}\vspace{1em}}
\chapter{Acknowledgements}

I wish to thank, first and foremost, my excellent PhD supervisors: Christian Schwanenberger and Alexander Grohsjean. You gave me feedback when I needed it, left me space when I did not, and always had my back in our common (mis-)adventures in the wide and wonderful and sometimes wondrous world that is Experimental Particle Physics. You also gave me great (and honest) advice and support for my further career, for which I am extremely thankful. Working with you has been a pleasure.

I am also deeply grateful to the many colleagues, postdocs and other PhD students, who taught me the intricacies of CMS and provided technical and moral support: Afiq Anuar, for unleashing $A/H \rightarrow \ttbar$ in the first place, and for always distracting me with interesting physics; Evan Ranken, who for some reason agreed to help perform a measurement in only two months; Jonas R{\"u}benach and Dominic Stafford, for collectively sparing me from ROOT with their framework; again Jonas, for setting up the $A/H$ analysis and giving me an excess to interpret; and again Dominic, for tolerating me as an officemate. Thanks to all of you.

Furthermore, I would like to thank my additional collaborators and supervisors on the different projects I worked on: Maria Aldaya and Andreas Meyer for the \ttbar cross section in Run 3; Simone Amoroso for \bbfourl; as well as Georg Weiglein, Sven Heinemeyer, and the Biek{\"o}tter Cousins for $\mathrm{ALPs} \rightarrow \ttbar$. 

Finally, I wish to give a huge thanks to the DESY CMS group for being a great place to work.